LDpred2: better, faster, stronger

### 2020-10-01 paper/paper-ldpred2.tex
prepare for bioRxiv

[TODO: ALSO TALK ABOUT LDPRED2 PARAMS AND POSSIBLY FUTURE WORK OF IMPROVING THEM VIA E.G. FUNCTIONAL ANNOTATIONS?]

[TODO: TALK ABOUT THE WRONG H2 ESTIMATION FOR CAD IN DISCU? -> CF. TABLE + SOME FIGURES IN SUPPMAT]

@@

LDpred2 also extends the original LDpred model in two ways.
First, we provide a sparse option in LDpred2-grid which provides models that truly encourage sparsity, in contrast to LDpred1 which outputs very small non-zero effect sizes \cite[]{cecile2019polygenic}.
In practice, the sparse version of LDpred2 can drastically reduce the number of variants used in the PGS without impacting its predictive performance, as opposed to discarding the smallest effects after having run LDpred (as tested in \cite{bolli2019software}).
The second extension is LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, which therefore does not require any validation set to tune hyper-parameters from. 
LDpred2-auto is an attractive option in many applications, especially since we now also provide an LD reference that can be used directly. 
LDpred2-auto almost always performs equally well as LDpred2-grid in simulations as well as in real data applications, provided that some quality control is performed on summary statistics (see Methods section ``Quality control of summary statistics'').
This quality control aims at ensuring that effects are transferable from the external GWAS summary statistics to the data where PGS are computed.

@@

We provide an LD reference for European ancestry to be used by researchers who do not have access to a large dataset to be used as LD reference.
We use the 362,320 UKBB individuals as used here, with some further quality control based on allele frequencies.
Indeed, some large allele frequency mismatches have been reported recently in the UKBB when comparing to other datasets, which are apparently due to mismappings \cite[]{kunert2020allele}.
Using 503 European individuals from the 1000 Genomes (1000G) data, we remove variants with allele frequency differences between UKBB and 1000G at $p<10^{-5}$ (Figure \ref{fig:af-outliers}).
We also remove variants with minor allele frequencies less than 1\% in the 1000G or less than 0.5\% in the UKBB.
1,054,330 variants remain, which we use to compute the LD reference provided.
We also provide an example R script on how to use this LD reference provided.

@@

The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
A tutorial on the steps to run LDpred2 using some small example data is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.

>>>>

@@

LDpred2 also extends the original LDpred model in two ways.
First, we provide a sparse option in LDpred2-grid which provides models that truly encourage sparsity, in contrast to LDpred1 which outputs very small non-zero effect sizes \cite[]{cecile2019polygenic}.
In practice, the sparse version of LDpred2 can drastically reduce the number of variants used in the PGS without impacting its predictive performance, as opposed to discarding the smallest effects after having run LDpred (as tested in \cite{bolli2019software}).
The second extension is LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, which therefore does not require any validation set to tune hyper-parameters from. 
LDpred2-auto is an attractive option in many applications, especially since we now also provide an LD reference that can be used directly. 
LDpred2-auto almost always performs equally well as LDpred2-grid in simulations as well as in real data applications.

However, LDpred2-auto requires that some quality control is performed on summary statistics (see Methods section ``Quality control of summary statistics''). 
This quality control aims at ensuring that effects are transferable from the external GWAS summary statistics to the data where PGS are computed.
Note that this quality control is also beneficial for other methods such as lassosum (data not shown).
Even when this quality control is performed, LDpred2-auto can slightly underperform in some situations, e.g. for CAD and Asthma here (Figure \ref{fig:AUC-all}).
When looking more closely at the results for CAD, even though chains seemed to have converged (Figure S5), we can see that the heritability estimate from LDpred2-auto is off compared to the one from LD score regression (Table S5), and that its estimate of $p$ (\textasciitilde0.001) is off compared to the best performing $p$ in LDpred2-grid (\textasciitilde0.01 in figure S4).
One possible explanation for this discrepancy is that the summary statistics we use for CAD come from a meta-analysis of 48 small studies, with some of them from non-European ancestries.
Moreover, the summary statistics we use for Asthma come from a meta-analysis of 66 small multiancestry studies.
This could break some of the assumptions used in LDpred2-auto and explain why LDpred2-auto underperforms for CAD and Asthma compared to LDpred2-grid.
Future work is needed to study how to make best use of these multiancestry meta-analyses from many small studies in the context of genetic prediction.
Future work is also needed to assess how relevant is the estimation of parameters from LDpred2-auto, and whether its estimates and its predictive performance could be improved by e.g. incorporating functional annotations.

@@

We provide an LD reference for European ancestry to be used by researchers who cannot compute their own.
We use the 362,320 UKBB individuals as used here, with some further quality control based on allele frequencies.
Indeed, some large allele frequency mismatches have been reported recently in the UKBB when comparing to other datasets, which are apparently due to mismappings \cite[]{kunert2020allele}.
Note that most of these errors should be captured by the quality control we propose in this paper.
Using 503 European individuals from the 1000 Genomes (1000G) data, we remove variants with allele frequency differences between UKBB and 1000G at $p<10^{-5}$ (Figure \ref{fig:af-outliers}).
We also remove variants with minor allele frequencies less than 1\% in the 1000G or less than 0.5\% in the UKBB.
1,054,330 variants remain, which we use to compute the LD reference provided.
We also provide an example R script on how to use this LD reference provided.

@@

The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
A tutorial on the steps to run LDpred2 using some small example data is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.
The European LD reference we provide here is available at \url{https://doi.org/10.6084/m9.figshare.13034123}.
